Copulas and their potential for ecology

### 2019-08-23 Paper.Rmd
edited the Intro to address referee concerns

# Abstract

All branches of ecology study relationships among environmental and biological variables. However, 
standard approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but not in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is exceedingly common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more 
strongly related in their left compared to right tails, or *vice versa*. We describe mechanisms by which 
observed copula structure and tail dependence can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. Copula structure and tail 
dependence can also help reveal causal relationships between variables. We also describe consequences of copula structure for ecological phenomena, including impacts 
on extinction risk and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to elevate copula reasoning to widespread use in ecology, benefiting the discipline.

@@

All branches of ecology 
study relationships among environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited, and 
provide only a partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may have the same correlation coefficient
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of dependence that can
have significance, as discussed below, but is not captured by 
correlation. 
Monotonic transformations can alter the Pearson, but not the Spearman
correlation (Appendix \ref{SM-pedagog1_details}, 
Fig. \ref{SM-fig_transf_pearson_spearman}). Although correlation
is only one way to 
study relationships between variables, other common metrics also 
provide partial information.

@@

<!--***DANr: reword this para so people don't think the ranks *are* the 
copulas, and to try to pre-defend against some of what ref 1 said. Also,
state that these are conceptual introductions only and are not formal,
a more comprehensive introduction is in xyz-->
Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The approaches are based on 
the concept of the *copula*, which we here 
introduce. Copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the dependence
structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the copula structures of 
A-B, respectively. Ranking makes the marginal distributions of 
the underlying random variables uniform, isolating the 
dependence structure.
<!--***DANr: Maybe move the below two sentences into the new section on Background-->
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.

A main benefit of a copula approach is 
that it can detect tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, D). Datasets of the same 
correlation can have a range of tail dependence 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for Q1 below).

The goal of this paper is to explore the potential for 
applications of copulas in ecology. 
<!--***DANr: Maybe emphasize here that we provide some, but not a comprehensive,
intro to copulas - for that go elsewhere. Instead we are trying to determine
to what extent they may be useful in ecology.-->
Our results provide support for the hypothesis that environmental, ecological and evolutionary processes 
generate and transmit nontrivial 
<!--***DANr: nontrivial? what does it mean-->
copula structure and tail dependence, with substantial 
consequences for theoretical and conceptual understanding, for inference, and for 
applications across ecology. The work we present here has begun to test this
hypothesis, and we hope our results stimulate further
evaluation of the hypothesis by the research community.

@@

<!--***DANr: Maybe use the words tail dependence more and the word copula
less in the below? Generally the words "copula structure" may be rubbing some
people wrong. Replacement words? What do you actually mean here? Maybe call it
asymmetric dependence, or complex dependence structure?-->
One way we may expect, *a priori*, a copula approach 
to better reflect ecological relationships relates to Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus soil N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association is visible in the copula
approach, but may not be revealed by standard techniques.

A second reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect an environmental variable 
measured through time in two locations to show strong tail dependence between
the locations if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case: extreme values are associated with intense events,
so happen in both places at the same time, whereas moderate values of the 
environmental variable are associated with local phenomena, which differ
between the locations.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This influence is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of the 
spatial synchrony of population, community, 
biogeochemical and other environmentally influenced ecological variables.
Synchrony is a phenomenon of substantial interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

<!--***DANr: Need some caveating here to accommodate the people who slavishly
follow the old maxim that correlation is not causation.-->
A third reason we may expect a copula approach to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
would suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F suggests that Sp2
is the dominant competitor. The copula approach again reveals which portions of the
distributions are principally responsible for the association between 
variables, and this 
information can then help support or refute alternative causal hypotheses
about which species is driving the dynamics of the other.

A fourth reason we may expect a copula approach to be important
for ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, captured by copulas, of the dependence between 
fluctuations of local
quantities can influence fluctuations of 
the spatial mean or sum, and how this may influence 
higher organizational levels in ecology and human concerns.
Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may be subject to a similar 
effect, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand left-tail dependence in a pest species, even if overall 
correlation were unchanged, would have more benign effects.

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have copula structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
dependence distinct from that of a normal distribution, and in particular
do they show asymmetric tail dependence? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. Such 
approaches do not utilize the power of copulas, so Q1 asks whether ecological
data contain dependency information distinct from what can be modelled with
standard approaches. (Q2) If the answer to
Q1 is "yes," what are some possible causes/mechanisms of non-normal
copula structure and tail dependence? And (Q3) what are the consequences of non-normal copula
structure and tail dependence for our understanding of ecology and for 
applications? We address Q1, in the second through fourth sections of the paper, 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2, in the fifth and sixth sections of the paper, using 
simple models. We address Q3, in the seventh and eigths sections of the paper, using 
both data and models. The ninth section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, non-normal copulas have not been used in ecology. 
We argue that benefits to ecology of using a complete 
copula description of dependence will be substantial.

>>>>

# Abstract
<!--***DANr: I have not edited this abstract at all since Ecol Monogr-->
All branches of ecology study relationships among environmental and biological variables. However, 
standard approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but not in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is exceedingly common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more 
strongly related in their left compared to right tails, or *vice versa*. We describe mechanisms by which 
observed copula structure and tail dependence can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. Copula structure and tail 
dependence can also help reveal causal relationships between variables. We also describe consequences of copula structure for ecological phenomena, including impacts 
on extinction risk and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to elevate copula reasoning to widespread use in ecology, benefiting the discipline.

@@

All branches of ecology 
study relationships among environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited, and 
provide only a partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may have the same correlation coefficient
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of dependence that can
have ecological significance, as discussed below, but is not captured by 
correlation. Although correlation is only one way to 
study relationships between variables, other common metrics also 
provide only partial information.

@@

<!--DONE ***DANr: reword this para so people don't think the ranks *are* the 
copulas, and to try to pre-defend against some of what ref 1 said. Also,
state that these are conceptual introductions only and are not formal,
a more comprehensive introduction is in xyz-->
Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The approaches are based on 
the concept of the *copula*; copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the section on Background below. Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cummulative distribution function (cdf) of the $x_i$ and $y_i$,
respectively. See, for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D}
relate to copulas in a way to be described in the section on Background below, 
after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals. Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables; 
they provide information solely on the dependence between those variables.

A main benefit of a copula approach is 
that it can detect tail dependence and asymmetries of tail dependence. 
Tail dependence (defined precisely in the section on Background, below) 
is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}D). Datasets of the same 
Spearman or Kendall correlation can have a range of tail dependence 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for Q1 below). Although copulas can be used to model the entire
complex dependence structure of data, we have found tail dependence to be
a useful aspect of the full structure of dependence, so tail dependence 
is one of the main foci of this paper.

The goal of this paper is to explore the potential for 
applications of copulas in ecology. We provide a sketch of an introduction
to copula concepts in the Background section below, but this is a sketch
only - readers can find comprehensive introductions elsewhere
[@nelsen2006_copula; @joe2014_dependence]. Instead our goal is to 
help determine to what extent ecological understanding may benefit from
using copulas to help understand complex dependence 
structures between variables.
Our results provide support for the hypothesis that environmental, 
ecological and evolutionary processes 
generate and transmit complex dependence structures, including tail dependence, 
with consequences for theory and 
applications. We hope our results will stimulate further
evaluation of this hypothesis by other researchers.

@@

<!--DONE ***DANr: Maybe use the words tail dependence more and the word copula
less in the below few paras? Generally the words "copula structure" may be rubbing some
people wrong. Replacement words? What do you actually mean here? Maybe call it
asymmetric dependence, or complex dependence structure?-->
One way we may expect, *a priori*, a copula 
study of complex dependence structures between
ecological variables
to better reflect ecological relationships relates to Liebig's law of the
minimum (the idea that growth is controlled not by total resources but by the scarcest 
resources). If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus soil N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association is visible in rank-based
plot, but may not be revealed by standard correlation techniques.

A second reason we may expect understanding of complex dependence structures
to benefit ecological
research is that prior work has demonstrated (using copulas) complex 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect an environmental variable 
measured through time in two locations to show strong tail dependence between
the locations if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case: extreme values are associated with intense events,
so happen in both places at the same time, whereas moderate values of the 
environmental variable may be associated with local phenomena, which differ
between the locations.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This influence is called the \emph{Moran effect}. 
If complex dependence structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect complex dependence 
structure to be a common feature of the 
spatial synchrony of population, community, 
biogeochemical and other environmentally influenced ecological variables.
Synchrony is a phenomenon of substantial interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

<!--DONE ***DANr: Need some caveating here to accommodate the people who slavishly
follow the old maxim that correlation is not causation.-->
A third reason we may expect an understanding of complex 
dependence structures to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
may suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F may suggest that Sp2
is the dominant competitor. Although other causal hypotheses may produce 
similar dependence structure, and it is usually impossible to obtain
complete information on causal pathways from analyses which are
fundamentally based on association data, more detailed 
examination of the patterns of dependence of variables may
well be combined with biological and other reasoning to rule out
some causal hypotheses which could not be eliminated solely 
with standard correlation approaches.

A fourth reason we may expect a copula understanding of complex 
dependence structures to be useful
for ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, which can be captured using copulas, 
of the dependence between fluctuations of local
quantities can influence fluctuations of 
the spatial mean or sum, and how this may influence 
higher organizational levels in ecology and human concerns.
Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may be subject to a similar 
effect, with
possible consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand left-tail dependence in a pest species, even if overall 
correlation were unchanged, would have more benign effects.

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
dependence distinct from that of a normal distribution, and in particular
do they show asymmetric tail dependence? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from what can be modelled with
standard approaches. (Q2) If the answer to
Q1 is "yes," what are some possible causes/mechanisms of non-normal
copula structure and tail dependence? And (Q3) what are the 
consequences of non-normal copula
structure and tail dependence for ecological understanding and
applications? After a Background section (the second section), we address Q1 
(third through fifth sections of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sixth and seventh sections of the paper), using 
simple models. We address Q3 (eighth and ninth sections of the paper) using 
both data and models. The final section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
but have been used only rarely in ecology. 
We argue that benefits of using a complete 
copula description of dependence in ecology will be substantial.

<!--***DANr: Maybe move the below two sentences into the new section on Background-->
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.
